\chapter*{Abstract}

This study proposes a new method 

\begin{note}

  
[X} is an essential part of any 

The problem ...

This study presents and prototypes a novel method which ....
That is based on visual programming, and static web applications. 

[How]

Using this, ...

This method turned out to indeed provide solve

The significance of this method lies herein that it ...

\end{note}

% Geocomputation is a cornerstone of GIS \& modern mapping needs.
% While the last decades have seen major advancements, writing a geocomputational pipeline remains a complex, non-trivial exercise. 
% Performing geocomputations by utilizing a visual programming language (VPL) within a web browser is a novel development which seeks to simplify this process, allowing more people to write these types of pipelines more successfully.
% It would, in theory, allow end-users to author geocomputation pipelines, without the need of a software development background, and without the need to install software.
% utilizing a dataflow model.
% This model is the standard within VPLs dealing with geometry processing, and contains many advantages in terms of performance and reliability.

In this context, the portability of existing geocomputation libraries is a common problem.
This is often addressed by maintaining duplicate JavaScript alternatives to libraries such as GDAL, PROJ, and GEOS, complicating innovation.  

This study seeks to improve the state of browser-based geocomputation VPLs by attempting to bring industry-standard geo-libraries to these environments. 
The study poses that this can only be done if native geocomputation libraries can be \emph{compiled} to WebAssembly, \emph{loaded} into a VPL, and \emph{utilized} in a browser-based dataflow-VPL format.
Discovering if and how these steps can be performed is the central question of this study. 
% This study seeks to improve the state of web-based geocomputation VPLs 
% by discovering if the library portability problem can be overcome, and if so, how.

This question is answered by testing a possible solution.
A proof of concept web-based VPL is designed, implemented and evaluated, which makes use of a novel plugin system, as well as a directed, acyclic graph-based data model \& interface.

Using this proof of concept, the study was able to demonstrate that the web platform was sufficiently capable of representing a dataflow VPL capable of constructing geocomputation pipelines.
The functional programming-properties of this dataflow VPL also makes geo-libraries sufficiently \emph{usable}, albeit with some well-known caveats of dataflow-VPLs, like the representation of conditionals and iteration.  

The current methods of \emph{compiling} existing C++ geocomputation libraries to the web turned out to be insufficient for the purposes of this study.  
This is due to the Emscripten compiler's focus on compiling full C++ applications instead of separate libraries. 
Despite this, the study wás able to demonstrate how a novel method can be used to sufficiently \emph{compile} and \emph{load} multiple Rust-libraries for usage in the VPL, thanks to more feature-rich WebAssembly tools in the Rust ecosystem. 
While Rusts geocomputation libraries are young, the study presents this method to either offer emscripten contributors a blueprint of a desired workflow, or to offer geocomputation library contributors a powerful use-case for the Rust language. 

All in all, this means that either if the geocomputation libraries found in the Rust ecosystem mature, or if Emscripten's capabilities improve, then the code portability problem \& dataflow problem of existing web-based geocomputation VPLS can be solved. 


% \begin{note}

%     "make a simple narrative of what you did and what are the results."
%     "make it not be a very rough draft"
%     "6.2: deliver the things you promised."
%     "make it more formal: replace: "crucial", "by far" or "impressive""

%     Stelios Draft Comments
%     ======================
    
%     In general, I think it's going well. I think the intro is fine, although it can be improved. I like the conclusions quite a lot, tbh. You are giving proper and interesting answers to the raised questions. The rest is a bit of a very rough draft, as I see. So, some specific points:
    
%     TODO: Tone down stuff
%     -  It seems like you have placed several bits here and there that seem like either introduction or conclusions. I feel your enthusiasm and excitement in your writing, but I think you are bringing questions and answers in places were people would normally just expect a simple narrative of what you did and what are the results.
    
%     TODO: Write 6.2
%     -  Speaking of results, we are missing the experiments so please put this at your first priority. This is because it feels like you promise things that you do not deliver.
    
%     TODO: Find & Replace
%     -  In general, you are using some very bold statements and informal expressions. The latter is kinda okay, but the first one is very dangerous. I've noticed words like "crucial", "by far" or "impressive" which seem very biased and subjective, so better be avoided. You can still use them here and there (mostly in the conclusions), but I think you use them way  too much (again, I can see your enthusiasm).
    
%     TODO: Find & Replace 
%     -  Minor thing, but I think you have confused \citet with \citep. The first one is used when the citation is part of the sentence and the latter when it's at the end of a sentence (or in a parenthesis, anyway).
    
% \end{note}

% This might allow web based VPLs for geocomputation to be 
% Geodata computation is important.
% Geodata computation is difficult.
% geo-web-vpls could help, but have seen little research
% This study: design, implement, analyze a new prototype geo-web-vpl. 
% Design-> utilize existing, native libraries written in C++ \& Rust on the web, in the format of a vpl.

% results -> it works.

% -> study shows that interplay between textual and visual programming is possible


% \begin{note}
%  - Performance intensive: (Big data, O(n^2) problems) 
%  - Heterogenous data (type, quality, scale, criteria, crs) 
%  - Complex (geometric) operations (linalg, bundle adjustment, procedural modelling) 
% \end{note}

% All of this makes the process of geocomputation difficult. 


% The full flowchart runs client-side in a browser, and both end results and intermediate products can be inspected in a 3D viewer.

% GeoFront offers several functionalities such as the parametric creation of 2D and 3D primitives, triangulation, isocurve extraction, and more. 
% These functionalities can be expanded upon though a plugin system which utilizes the existing "Node Package Manager" infrastructure.
% Together with WebAssembly, this enables the utilization of industry standard geoprocessing libraries such as `CGAL`, `GDAL` and `PROJ`, and data parsing libraries such as `IFC.js` and `laz-rs`.



% Following the implementation, the project was tested by simulating use-case scenarios. 
% The tests demonstrate the feasibility of [...]
% At the same time some key parameters of [...] identified which if tuned properly they can optimize the performance, behavior and robustness of the geo-web-vpl.
% With the project being a prototype solution, the vpl is far from operational and there is certainly a lot of space for improvement regarding both components. 

% 1. Tryout (ACTUAL)
%    - A-la wapm WebAssembly Package Manager allows packages to be run from within the package-page itself. 
%   - Just meant to quickly try out some features.

% 2. Educational (ACTUAL)
%    - interactive educational tool
%    - (What does a delaunay triangulation look like? how does it behave? What happens if you lower the radius of inverse distance weighting ? )

% 3. Rapid-Prototyping (POSSIBLE)
%    - Setting up pipelines which can be consumed by cloud-based geoprocessing services. 
%    - Future work: export flowchart to a process which can be run natively or server side.

% 4. Publishing (POSSIBLE)
%    - Geotiff.io / ModelLab
%    - Web FME 
%    - Publish full web apps in and off themselves, making use of zero, one or multiple wasm-compiled libraries.  
%    - Future work: export to web-app (without flowchart)

% % Safesoft's FME, but web based \& open source 

% % CONCLUSION
% By creating geofront, this thesis was able to discover .............

% - The web is able to facilitate a visual proregramming language.

% - The web is able to be used for geoprocessing, albeit with some caveats
%   - TypedArrays,
%   - Geometric predicates 
%   - Rounding
%   - ETC.

% - Many of these things can be fixed with webassembly, but webassembly itself has other shortcomings
%   - Differences between Rust \& C++

% - reasonable performance 
%   (- great considering the platform)

% - would not be possible without these modern web features
%   - Web Assembly 
%   - Typed Array's 
%   - Web Workers
%   - Web Components,
%   - 2D Canvas API
%   - Web GL

% \todo{[JF]: I need to add more critical notes on promises of accessibility. Is a webapp truly accessible? Is a flowchart interactive, or does it hinder interactiveness? }

% I believe that such a web application can make geoprocessing more accessible to practitioners.
% This empowers users to create small geoprocessing demo's, and share these 
% With geofront, geoprocessing libraries can be loaded and used interactively. Users are also able to create and share flowcharts.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% 


%%%%%



\begin{note}

  Lessons
  ==========
  
  - give yourself clear todo's per day. 
    Not too little, not too much.
    Upon completion, the day falls apart
  
  
  TODO
  ============
  
  high-level
  ----------
  
  [x] The intro / story needs to be altered 
      - stronger tie to Geo & web
      - VPL + Web
  
  [ ] rewrite and balance research questions
  - [ ] than trickle down the consequences of those changes throughout the thesis
  - [ ] and fix all that needs to be fixed (images and stuff)
  - [ ] put TODO notes everywhere
  mid-level
  ---------
  
  [ ] TODO: rewrite abstract in accordance with new story

  [x] Deep dive in Hugo's comments, fix those aspects 
    - [x] sort acronyms
  
  [x] Add a stronger web-component in the introduction (presentation)
  [x] Add cloud native / cloud-based geodata component to the introduction  
  
  [ ] Cleanup Chapter 2 and 3.  
    - Some parts are still more relevant than other parts. 
    - Also, there are some things I say or use in chapter 4, 5, and 6, that were not properly explained in chapter 2. Examples of this are emscripten, or CGAL. 
    - However, a complete rewrite would be too much work.
    - Would love to hear your take on this.
  
  [ ] Make chapter 4 and 5 more streamlined 
    - I've noticed that I repeat myself quite often.
    - I think this is because I was not sure if I already said something relevant in the chapter before.
  
  [ ] Show your results more clearly, do the coding you have done justice
  
  [ ] finish chapter 6 properly 
    - Experiments: add a section on a use-case application, using potree + startin + geofront std + obj downloader
  
  [ ] write stronger, more clear conclusions.
    - hmm, i think they are already quite clear. But again, this is to be reconsidered when the overall story of the thesis changes  
  
  low-level: thursday & friday
  -------------------

  - [ ] TODO: write Acknowledgements
  - [x] Add arsclassica template to the thesis
  - [x] Make the code snippets / listings work properly
  - [x] replace all code screenshot with proper listings
  - [ ] create all missing uml diagrams
  - [ ] add proper graphs for the data you've gathered 
  - [ ] Add all proper metadata to Zotero, for a nice bibliography
  - [ ] Fix all TODO graphics
  - [ ] Fix all acronyms,
  - [ ] Sort the acronyms at the end 
  - [ ] Make sure CiteT and citeP is properly used everywhere. No more raw 'et. al.' statments 
  - [ ] Capitalize all VPL / replace with acronyms
  - [ ] Sources need to be fixed (more data, check if I can do things like 'empscripten contributors')
  - [ ] Replace C++ into C / C++ everywhere, but especially the intro
\end{note}
