
\chapter*{Acknowledgements}

The completion of this thesis would not have been possible without the generous support of the people around me. 
The least I can offer back is my most sincere gratitude.
First of all, I want to thank Ken Arroyo Ohori and Stelios Vitalis for going above and beyond in supporting me during all phases of this thesis.
They have invested many personal hours along the full scope of this journey, and I couldn't have asked for better guidance. 
Thank you Giorgio Agugiaro, for seamlessly stepping in during the later stages of the process, and for steering me towards finalization. 
Thank you Hugo Ledoux for the crucial suggestions and comments regarding the \emph{cloud-native} aspect of this study.

My employers and colleagues during the execution of this thesis also played a vital role. 
Thank you Sybren de Graaf, my previous employer at Sfered, for your interest, and the opportunity to develop visual programming, which greatly influenced the subject of this thesis.
I want to thank Martin Kodde, my current employer at Geodelta, for the opportunity to learn geodesy in practice, and my colleagues there for their continued interest. 
In particular, thank you Roeland Boeters, for hosting the dataset used in the Potree demo.

I want to thank my fellow students Laurens, Maarit, Mels, Siebren, Alex \& Max, for all the fun and interesting conversations we have had during our studies. 
My friends from D.S.J.V. Groover also greatly supported me. In particular I want to thank Tim Boot for proofreading my P2. I also cannot forget my mom, dad, brother, and the rest of the family for their encouragement and pride. 

Finally, and most importantly, I would like to thank Nadja for her patience and unwavering support throughout the entire period.

Thank you all!

